---
title: "Charts and Data Visualisation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Charts and Data Visualisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

rtui includes 12 chart types rendered directly in the terminal using
[plotext](https://github.com/piccolomo/plotext) via
[textual-plotext](https://github.com/Textualize/textual-plotext), plus the
ability to render ggplot2 objects as terminal charts.

> **Terminal only:** All examples must be saved as `.R` files and run from a
> real terminal (`Rscript my_app.R`). rtui apps do **not** work in RStudio,
> R GUI, Jupyter, or any embedded R console.

## Setting up a chart

Charts are drawn on `text_plot()` widgets. First, add a `text_plot()` to your
layout, then call `plot_*()` functions from handlers to draw data:

```{r}
library(rtui)

quick_app(
  title = "My Chart",
  layout = vstack(
    header(),
    text_plot(id = "chart"),
    footer()
  ),
  on_mount = function(event, state) {
    plot_bar(state$app, "chart",
             labels = c("A", "B", "C", "D"),
             values = c(25, 40, 15, 30),
             title = "Sales by Region",
             color = "cyan")
    state
  },
  bindings = list(binding("q", "quit_app", "Quit", priority = TRUE)),
  on_action = function(event, state) {
    if (event$value == "quit_app") return(quit())
    state
  },
  css = "#chart { height: 1fr; }"
)
```

## Chart types

### Bar chart

```{r}
plot_bar(state$app, "chart",
         labels = c("Mon", "Tue", "Wed", "Thu", "Fri"),
         values = c(120, 200, 150, 180, 220),
         title = "Daily Visitors",
         color = "green",
         xlabel = "Day",
         ylabel = "Visitors")

# Horizontal orientation
plot_bar(state$app, "chart",
         labels = c("R", "Python", "Julia"),
         values = c(80, 95, 40),
         orientation = "horizontal",
         color = "blue")
```

### Line chart

```{r}
x <- 1:50
y <- cumsum(rnorm(50))

plot_line(state$app, "chart",
          x = x, y = y,
          title = "Random Walk",
          color = "cyan",
          marker = "braille",
          xlabel = "Step",
          ylabel = "Value")
```

The `marker` parameter controls the drawing style: `"braille"` (highest
resolution), `"dot"`, `"hd"`, `"fhd"`, or standard characters.

### Scatter plot

```{r}
plot_scatter(state$app, "chart",
             x = mtcars$wt,
             y = mtcars$mpg,
             title = "Weight vs. MPG",
             color = "magenta",
             marker = "braille",
             xlabel = "Weight (1000 lbs)",
             ylabel = "Miles per Gallon")
```

### Histogram

```{r}
plot_hist(state$app, "chart",
          data = rnorm(1000),
          bins = 20,
          title = "Normal Distribution",
          color = "yellow",
          xlabel = "Value",
          ylabel = "Frequency")
```

### Box plot

Compare distributions across groups:

```{r}
plot_box(state$app, "chart",
         data = list(
           "4 cyl" = mtcars$mpg[mtcars$cyl == 4],
           "6 cyl" = mtcars$mpg[mtcars$cyl == 6],
           "8 cyl" = mtcars$mpg[mtcars$cyl == 8]
         ),
         title = "MPG by Cylinders",
         xlabel = "Cylinders",
         ylabel = "MPG")
```

### Stacked bar chart

Show composition across categories:

```{r}
plot_stacked_bar(state$app, "chart",
                 labels = c("Q1", "Q2", "Q3", "Q4"),
                 data = list(
                   "Product A" = c(10, 15, 12, 18),
                   "Product B" = c(8, 12, 15, 10),
                   "Product C" = c(5, 8, 7, 12)
                 ),
                 title = "Quarterly Revenue by Product",
                 colors = c("red", "green", "blue"))
```

### Grouped bar chart

Side-by-side comparison:

```{r}
plot_multiple_bar(state$app, "chart",
                  labels = c("2022", "2023", "2024"),
                  data = list(
                    Revenue = c(100, 150, 200),
                    Expenses = c(80, 110, 130),
                    Profit = c(20, 40, 70)
                  ),
                  title = "Financial Overview",
                  colors = c("green", "red", "cyan"))
```

### Heatmap

Visualise matrix data as colour intensity:

```{r}
mat <- matrix(runif(50), nrow = 5, ncol = 10)

plot_heatmap(state$app, "chart",
             matrix = mat,
             title = "Activity Heatmap",
             xlabel = "Hour",
             ylabel = "Day")

# Works with data.frames too
plot_heatmap(state$app, "chart",
             matrix = cor(mtcars[, 1:6]),
             title = "Correlation Matrix")
```

### Candlestick chart

Financial OHLC data:

```{r}
dates <- format(Sys.Date() - 9:0, "%m/%d")

plot_candlestick(state$app, "chart",
                 dates = dates,
                 data = list(
                   open  = c(100, 102, 101, 105, 103,
                             107, 106, 110, 108, 112),
                   high  = c(104, 105, 106, 108, 107,
                             110, 111, 114, 112, 115),
                   low   = c(98, 100, 99, 103, 101,
                             105, 104, 108, 106, 110),
                   close = c(102, 101, 105, 103, 107,
                             106, 110, 108, 112, 114)
                 ),
                 title = "ACME Corp Stock",
                 colors = c("green", "red"))
```

### Error bars

```{r}
x <- 1:5
y <- c(10, 20, 15, 25, 18)
yerr <- c(2, 3, 1, 4, 2)

plot_error(state$app, "chart",
           x = x, y = y, yerr = yerr,
           title = "Measurements with Error",
           color = "cyan")
```

### Event plot

Mark events along a timeline:

```{r}
events <- c(5, 12, 23, 35, 42)

plot_event(state$app, "chart",
           positions = events,
           title = "System Events",
           color = "red",
           xlabel = "Time (minutes)")
```

## Chart options

All `plot_*()` functions share common parameters:

| Parameter | Description |
|-----------|-------------|
| `title` | Chart title displayed above the plot |
| `color` / `colors` | Colour name(s): `"red"`, `"green"`, `"cyan"`, etc. |
| `xlabel` | X-axis label |
| `ylabel` | Y-axis label |
| `clear` | Clear previous plot data first (default `TRUE`) |
| `orientation` | `"vertical"` or `"horizontal"` (bar/box charts) |

## Layering multiple series

Set `clear = FALSE` to overlay data on an existing plot:

```{r}
on_mount = function(event, state) {
  x <- 1:100

  # First series (clears canvas)
  plot_line(state$app, "chart",
            x = x, y = sin(x / 10),
            title = "Trigonometric Functions",
            color = "cyan", marker = "braille",
            clear = TRUE)

  # Second series (overlay)
  plot_line(state$app, "chart",
            x = x, y = cos(x / 10),
            color = "magenta", marker = "braille",
            clear = FALSE)

  # Third series
  plot_line(state$app, "chart",
            x = x, y = sin(x / 10) * cos(x / 20),
            color = "yellow", marker = "braille",
            clear = FALSE)

  state
}
```

## Live-updating charts

Combine charts with timers for real-time visualisation:

```{r}
library(rtui)

quick_app(
  title = "Live Chart",
  layout = vstack(
    header(),
    text_plot(id = "chart"),
    sparkline(c(0), id = "spark"),
    footer()
  ),

  on_mount = function(event, state) {
    state$set("values", numeric(0))
    set_interval(state$app, 0.5, "update_chart")
    state
  },

  on_timer = function(event, state) {
    if (event$timer_id == "update_chart") {
      vals <- c(state$get("values"), runif(1, 0, 100))
      if (length(vals) > 50) vals <- tail(vals, 50)
      state$set("values", vals)

      plot_line(state$app, "chart",
                x = seq_along(vals), y = vals,
                title = "Live Sensor Data",
                color = "cyan", marker = "braille")

      update(state$app, "spark", data = vals)
    }
    state
  },

  bindings = list(binding("q", "quit_app", "Quit", priority = TRUE)),
  on_action = function(event, state) {
    if (event$value == "quit_app") return(quit())
    state
  },
  css = "
    #chart { height: 1fr; }
    #spark { height: 3; }
  "
)
```

## Rendering ggplot2 objects

If you use ggplot2, `plot_ggplot()` can render your plots directly in the
terminal. It extracts layer data from the ggplot object and maps each geom
to the appropriate terminal chart type.

```{r}
library(rtui)
library(ggplot2)

quick_app(
  title = "ggplot in Terminal",
  layout = vstack(
    header(),
    text_plot(id = "chart"),
    footer()
  ),

  on_mount = function(event, state) {
    p <- ggplot(mtcars, aes(wt, mpg)) +
      geom_point() +
      geom_smooth(method = "lm") +
      labs(title = "Weight vs. MPG",
           x = "Weight (1000 lbs)",
           y = "Miles per Gallon")

    plot_ggplot(state$app, "chart", p)
    state
  },

  bindings = list(binding("q", "quit_app", "Quit", priority = TRUE)),
  on_action = function(event, state) {
    if (event$value == "quit_app") return(quit())
    state
  },
  css = "#chart { height: 1fr; }"
)
```

### Supported geoms

`plot_ggplot()` handles these ggplot2 geom types:

| ggplot2 geom | Terminal chart type |
|---|---|
| `geom_point()`, `geom_jitter()` | Scatter |
| `geom_line()`, `geom_path()`, `geom_smooth()`, `geom_step()` | Line |
| `geom_col()`, `geom_bar()` | Bar |
| `geom_histogram()`, `geom_freqpoly()` | Bar (pre-binned) |
| `geom_boxplot()` | Box |
| Other geoms with x/y data | Falls back to scatter |

Labels and titles from ggplot's `labs()` are preserved.

### ggplot2 gallery example

Cycle through multiple ggplot2 charts using tabs:

```{r}
library(rtui)
library(ggplot2)

plots <- list(
  scatter = ggplot(mtcars, aes(wt, mpg)) + geom_point() +
    labs(title = "Scatter: Weight vs MPG"),
  histogram = ggplot(mtcars, aes(mpg)) + geom_histogram(bins = 10) +
    labs(title = "Histogram: MPG Distribution"),
  bar = ggplot(mtcars, aes(factor(cyl))) + geom_bar() +
    labs(title = "Bar: Cars by Cylinder Count"),
  box = ggplot(mtcars, aes(factor(cyl), mpg)) + geom_boxplot() +
    labs(title = "Box: MPG by Cylinders")
)

quick_app(
  title = "ggplot Gallery",
  layout = vstack(
    header(),
    tabs(
      tab_pane("Scatter", text_plot(id = "p1"), title = "Scatter", id = "t1"),
      tab_pane("Histogram", text_plot(id = "p2"), title = "Histogram", id = "t2"),
      tab_pane("Bar", text_plot(id = "p3"), title = "Bar", id = "t3"),
      tab_pane("Box", text_plot(id = "p4"), title = "Box", id = "t4"),
      id = "chart_tabs"
    ),
    footer()
  ),

  on_mount = function(event, state) {
    set_timer(state$app, 0.2, "draw_plots")
    state
  },

  on_timer = function(event, state) {
    if (event$timer_id == "draw_plots") {
      plot_ggplot(state$app, "p1", plots$scatter)
      plot_ggplot(state$app, "p2", plots$histogram)
      plot_ggplot(state$app, "p3", plots$bar)
      plot_ggplot(state$app, "p4", plots$box)
    }
    state
  },

  bindings = list(binding("q", "quit_app", "Quit", priority = TRUE)),
  on_action = function(event, state) {
    if (event$value == "quit_app") return(quit())
    state
  },
  css = paste0(tui_theme("monokai"), "
    TextPlot { height: 1fr; }
  ")
)
```

## Data tables

For tabular data, `data_table()` provides a full interactive table with
sorting and selection:

```{r}
quick_app(
  title = "Data Explorer",
  layout = vstack(
    header(),
    data_table(
      iris,
      id = "iris_table",
      cursor = "row",
      zebra_stripes = TRUE,
      sortable = TRUE
    ),
    footer()
  ),
  bindings = list(binding("q", "quit_app", "Quit", priority = TRUE)),
  on_action = function(event, state) {
    if (event$value == "quit_app") return(quit())
    state
  }
)
```

Or use the one-liner convenience function:

```{r}
data_viewer(iris)
```

### Dynamic table updates

Add rows, clear, or rebuild tables at runtime:

```{r}
# Add new rows (pass as list)
new_rows <- data.frame(
  Sepal.Length = 5.0,
  Sepal.Width = 3.5,
  Petal.Length = 1.5,
  Petal.Width = 0.2,
  Species = "setosa"
)
update(state$app, "iris_table", add_rows = as.list(new_rows))

# Clear all data
update(state$app, "iris_table", clear_data = TRUE)
```

## Sparklines

Sparklines are compact, inline charts perfect for dashboards:

```{r}
sparkline(c(1, 4, 2, 7, 3, 8, 5, 9), id = "cpu_spark")

# Update with new data
update(state$app, "cpu_spark", data = new_values)
```

## Combining charts in a dashboard

```{r}
library(rtui)

quick_app(
  title = "Sales Dashboard",
  layout = vstack(
    header(),
    hstack(
      vstack(
        static("[bold]Revenue[/bold]"),
        text_plot(id = "revenue_chart"),
        id = "left_panel"
      ),
      vstack(
        static("[bold]Distribution[/bold]"),
        text_plot(id = "dist_chart"),
        id = "right_panel"
      )
    ),
    hstack(
      sparkline(c(0), id = "trend1"),
      sparkline(c(0), id = "trend2"),
      sparkline(c(0), id = "trend3")
    ),
    footer()
  ),

  on_mount = function(event, state) {
    # Revenue bar chart
    plot_bar(state$app, "revenue_chart",
             labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun"),
             values = c(42, 55, 38, 67, 72, 61),
             title = "Monthly Revenue ($K)",
             color = "green")

    # Distribution histogram
    plot_hist(state$app, "dist_chart",
              data = rlnorm(500, meanlog = 3, sdlog = 0.5),
              bins = 15,
              title = "Order Size Distribution",
              color = "cyan")

    # Sparkline trends
    update(state$app, "trend1", data = cumsum(rnorm(20)))
    update(state$app, "trend2", data = cumsum(rnorm(20)))
    update(state$app, "trend3", data = cumsum(rnorm(20)))

    state
  },

  bindings = list(binding("q", "quit_app", "Quit", priority = TRUE)),
  on_action = function(event, state) {
    if (event$value == "quit_app") return(quit())
    state
  },
  css = paste0(
    tui_theme("ocean"),
    "
    #left_panel, #right_panel { width: 1fr; padding: 0 1; }
    TextPlot { height: 1fr; }
    Sparkline { height: 3; margin: 0 2; }
    "
  )
)
```
