---
title: "Example Gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example Gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

rtui ships with 14 ready-to-run example apps in `inst/examples/`. Each one
is self-contained and demonstrates a progressively richer set of features.

> **Terminal only:** Save as `.R` files and run from a real terminal with
> `Rscript`. These will **not** work in RStudio or R GUI.

## Running examples

```r
# From inside the rtui project directory:
Rscript inst/examples/05-counter.R

# Or after installing the package:
system.file("examples", "05-counter.R", package = "rtui") |>
  source()
```

## 01 — Hello World

**Features:** `text()`, `box()`, `vstack()`, key handler, `quit()`.

The simplest possible rtui app — a styled greeting with a quit key.

```{r}
library(rtui)

app <- tui_app(
  layout = vstack(
    box(
      text("Hello from rtui!", id = "greeting"),
      border = "round", title = "Welcome"
    ),
    text("Press 'q' to quit.", id = "hint"),
    id = "root"
  ),
  on_key = function(event, state) {
    if (event$key == "q") return(quit(state))
    state
  }
)

app$run()
```

## 02 — List-Detail

**Features:** `list_view()`, `on_change`, `update()`, per-widget routing.

A master-detail pattern: select an item on the left, see its content on
the right.

## 03 — Data Table

**Features:** `data_table()`, displaying data frames.

Pass any data.frame directly to `data_table()` for an interactive,
scrollable table view.

## 04 — dfdiff Explorer

**Features:** `list_view()`, `static()`, `on_mount`, state timestamps.

A more complete list-detail explorer with named sections and mount-time
state initialization.

## 05 — Reactive Counter

**Features:** `reactive()` bindings, `digits()`, `button()`, click
handlers, `notify()`.

The classic counter app — but with reactive bindings so `state$set("count", n)`
automatically updates the digit display.

```{r}
library(rtui)

quick_app(
  title = "Counter",
  layout = vstack(
    header(),
    center(vstack(
      digits("0", id = "display"),
      hstack(
        button("-1", id = "dec"),
        button("Reset", id = "reset"),
        button("+1", id = "inc")
      )
    )),
    footer(),
    id = "root"
  ),
  reactive = reactive(count = "display"),
  on_mount = function(event, state) {
    state$set("count", 0L)
    state
  },
  on_click = list(
    inc = function(event, state) {
      state$set("count", state$get("count", 0L) + 1L)
      state
    },
    dec = function(event, state) {
      state$set("count", state$get("count", 0L) - 1L)
      state
    },
    reset = function(event, state) {
      state$set("count", 0L)
      notify(state$app, "Counter reset.", severity = "info")
      state
    }
  )
)
```

## 06 — Contact Form

**Features:** `tui_form()`, `input()`, `select()`, `checkbox()`,
`confirm()` dialog, `on_screen_result`.

A structured form with validation, confirmation dialog before submission,
and status feedback.

## 07 — Stopwatch

**Features:** `set_interval()`, `clear_timer()`, `progress_bar()`,
`log_view()`, key bindings with `binding()`.

Start/pause/reset stopwatch with lap tracking. Press Space to toggle,
`d` for dark mode.

## 08 — Tabbed Dashboard

**Features:** `tabs()`, `tab_pane()`, `data_table()`, `sparkline()`,
`plot_bar()`, `digits()`, `tui_theme()`.

A three-tab sales dashboard with KPI cards, sparkline trends, a sortable
data table, and a bar chart — styled with the Catppuccin theme.

## 09 — Todo List

**Features:** `option_list()`, `on_submit`, dynamic list updates,
`confirm()` dialog, CRUD operations.

Add tasks by typing and pressing Enter, mark them done, delete with
confirmation, and clear all completed items.

## 10 — Chart Gallery

**Features:** `plot_bar()`, `plot_line()`, `plot_scatter()`, `plot_hist()`,
`plot_heatmap()`, tabs, Dracula theme.

Five chart types in a tabbed view, each with sample data.

## 11 — Screens & Modals

**Features:** `push_screen()`, `pop_screen()`, `tui_screen()`,
`confirm()`, `alert()`, screen results.

A multi-screen app with a settings modal (input, select, switch) that
returns results to the main screen.

## 12 — Reactive Dashboard

**Features:** `reactive()` with formula bindings, `set_interval()`,
auto-updating sparklines, `log_view()`, multiple reactive targets, Nord
theme.

A simulated system monitor that updates CPU/memory/disk every 2 seconds.
Reactive formulas auto-update digits and progress bars. Warnings are
logged and notified.

```{r}
reactive = reactive(
  cpu = list(
    ~ update(.app, "cpu_display", value = paste0(.x, "%")),
    ~ update(.app, "cpu_bar", progress = .x)
  ),
  mem = list(
    ~ update(.app, "mem_display", value = paste0(.x, "%")),
    ~ update(.app, "mem_bar", progress = .x)
  )
)
```

## 13 — Background Tasks

**Features:** `run_async()`, `cancel_async()`, `on_task` handler,
`log_view()`, `progress_bar()`, error handling.

Launch background R processes without freezing the UI. Run a slow
computation, see results arrive asynchronously, or cancel a running task.

```{r}
# Run a background task
run_async(state$app, function() {
  Sys.sleep(3)
  42
}, name = "my_task")

# Handle completion
on_task = function(event, state) {
  if (event$widget_id == "__async_ok") {
    notify(state$app, paste("Result:", event$value))
  }
  state
}
```

## 14 — Hot Reload Development

**Features:** `dev_app()`, file watching, automatic restart.

A development tool that watches your `.R` file and restarts the TUI
whenever you save. Edit in your favourite editor, save, and see changes
instantly.

```r
# Launch from terminal:
Rscript -e "rtui::dev_app('my_app.R')"
```

## Showcase apps

Beyond these examples, rtui includes 8 full showcase apps in `spikes/`:

| App | Highlights |
|-----|------------|
| Stock Tracker | Live prices, candlestick charts, sparklines |
| System Monitor | CPU/memory/disk gauges, process table |
| Log Viewer | Severity filter, search, pause/resume |
| CSV Explorer | Browse R datasets, 5 chart types |
| ggplot Explorer | 8 ggplot2 plots rendered in terminal |
| Pomodoro Timer | Big-digit countdown, work/break cycles |
| Git Dashboard | Commit log, contributor charts |
| Markdown Notes | CRUD notes with live markdown preview |

Run any showcase:

```bash
Rscript spikes/showcase_pomodoro.R
```
