---
title: "Getting Started with rtui"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rtui}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What is rtui?

rtui lets you build full-screen terminal user interfaces (TUIs) from R. It
wraps Python's [Textual](https://textual.textualize.io/) framework via
[reticulate](https://rstudio.github.io/reticulate/), giving you access to 35+
widgets, 12 chart types, CSS-like styling, reactive state, screens, timers,
key bindings, a command palette, and 10 built-in colour themes -- all without
writing a single line of Python.

> **Terminal only:** rtui apps require a real terminal emulator (Windows
> Terminal, iTerm2, Terminal.app, or any xterm-compatible terminal). They
> will **not** work in the RStudio console, R GUI, Jupyter notebooks, or any
> embedded R console. Always save your code as a `.R` file and run it with
> `Rscript`.

## Installation

### Step 1: Install the R package

```{r}
# From a local checkout
devtools::install_local("path/to/rtui")

# Or from GitHub (when published)
# devtools::install_github("ariji/rtui")
```

### Step 2: Install Python dependencies

rtui needs Python >= 3.9 with the `textual` and `textual-plotext` packages.
A one-time setup function handles everything:
```{r}
rtui::install_python_deps()
```

This creates a virtualenv called `r-rtui` and installs the required Python
packages into it.

### Windows-specific setup

On Windows, avoid using the Microsoft Store version of Python. Instead, use
a standard Python install from [python.org](https://www.python.org/).

Add these lines at the top of your scripts (before `library(rtui)`):

```{r}
Sys.setenv(RETICULATE_PYTHON_ENV = "r-rtui")
library(reticulate)
use_virtualenv("r-rtui", required = TRUE)
```

### Requirements

- R >= 4.1
- Python >= 3.9
- A real terminal (not the RStudio console -- use Windows Terminal, iTerm2,
  Terminal.app, or any xterm-compatible terminal)

## Your first app

The fastest way to get started is `quick_app()`, which builds and runs an app
in a single call:

```{r}
library(rtui)

quick_app(
  title = "Hello",
  layout = center(
    middle(
      text("Hello from rtui!")
    )
  )
)
```

Run this from a real terminal:

```bash
Rscript my_app.R
```

You should see a full-screen app with "Hello from rtui!" centred on screen.
Press `Ctrl+C` to exit.

## One-liner data viewer

Want to explore a data frame interactively? One line:

```{r}
data_viewer(mtcars)
```

This opens a full-screen sortable data table. Click column headers to sort.
Press `q` to quit.

## A counter app

Let's build something interactive -- a counter with increment, decrement, and
reset buttons:

```{r}
library(rtui)

quick_app(
  title = "Counter",
  layout = vstack(
    header(),
    center(middle(vstack(
      digits("0", id = "count"),
      hstack(
        button("-1", id = "dec"),
        button("+1", id = "inc"),
        button("Reset", id = "reset")
      )
    ))),
    footer()
  ),

  on_click = list(
    inc = function(event, state) {
      n <- state$get("n", 0L) + 1L
      state$set("n", n)
      update(state$app, "count", value = as.character(n))
      state
    },
    dec = function(event, state) {
      n <- state$get("n", 0L) - 1L
      state$set("n", n)
      update(state$app, "count", value = as.character(n))
      state
    },
    reset = function(event, state) {
      state$set("n", 0L)
      update(state$app, "count", value = "0")
      state
    }
  ),

  bindings = list(
    binding("q", "quit_app", "Quit", priority = TRUE)
  ),
  on_action = function(event, state) {
    if (event$value == "quit_app") return(quit())
    state
  }
)
```

### What's happening here?

1. **Layout**: `vstack()` stacks children vertically. Inside we use `center()`
   and `middle()` to centre the content. `digits()` shows large-format numbers,
   and `button()` creates clickable buttons.

2. **Event handling**: `on_click` is a named list mapping widget ids to handler
   functions. Each handler receives `event` (with click details) and `state`
   (mutable key-value store). Handlers must return `state`.

3. **Updating widgets**: `update(state$app, "count", value = "5")` changes the
   digits widget's display. Every widget can be updated by id.

4. **Key bindings**: `binding("q", "quit_app", "Quit")` maps the `q` key to an
   action called `"quit_app"`. The `on_action` handler checks `event$value` and
   calls `quit()` to exit.

## The two-step pattern: tui_app() + run()

`quick_app()` is a convenience wrapper. For more control, use `tui_app()` to
create the app object, then call `$run()`:

```{r}
app <- tui_app(
  title = "My App",
  layout = vstack(
    header(),
    text("Built with tui_app()", id = "msg"),
    footer()
  ),
  on_key = function(event, state) {
    if (event$key == "q") return(quit())
    state
  }
)

# Inspect before running
print(app)

# Run it
app$run()
```

This is useful when you want to inspect or modify the app object before
launching.

## Running apps (terminal only!)

TUI apps take over the entire terminal screen. They will **not** work in:

- **RStudio** (neither the console nor the terminal pane)
- **R GUI** (Rgui.exe on Windows, R.app on macOS)
- **Jupyter notebooks**
- **Emacs ESS / VS Code R console** or any other embedded R console

You must save your code as a `.R` file and run it from a **real terminal
emulator**:

```bash
# macOS / Linux
Rscript my_app.R

# Windows (PowerShell)
& "C:\Program Files\R\R-4.4.3\bin\Rscript.exe" my_app.R
```

## Dark and light mode

Apps default to dark mode. Toggle at creation or at runtime:

```{r}
# Start in light mode
quick_app(title = "Light", dark = FALSE, layout = ...)

# Toggle at runtime from a handler
on_action = function(event, state) {
  if (event$value == "toggle_dark") dark_toggle(state$app)
  state
}
```

## Applying a theme

rtui ships with 10 colour themes. Pass one to the `css` parameter:

```{r}
quick_app(
  title = "Themed App",
  layout = vstack(header(), text("Dracula theme!"), footer()),
  css = tui_theme("dracula")
)
```

Available themes: `dracula`, `nord`, `monokai`, `solarized_dark`,
`solarized_light`, `gruvbox`, `catppuccin`, `ocean`, `forest`, `sunset`.

List them all:

```{r}
list_themes()
```

## Notifications

Show transient notification toasts from any handler:

```{r}
notify(state$app, "File saved!", severity = "info")
notify(state$app, "Disk full!", severity = "error")
```

Severity levels: `"info"`, `"warning"`, `"error"`.

## What's next?

Now that you have the basics, explore these guides:

- **[Layouts and Widgets](layouts-and-widgets.html)**: All containers and
  widgets with examples
- **[State, Events, and Reactivity](state-events-reactivity.html)**: Deep
  dive into handlers, state management, and reactive bindings
- **[Charts and Data Visualisation](charts-and-data.html)**: Terminal
  charts and ggplot2 rendering
- **[Advanced Features](advanced-features.html)**: Screens, timers, key
  bindings, command palette, forms, and workers
- **[Example Gallery](examples.html)**: 12 runnable example apps with
  progressive complexity
