---
title: "Layouts and Widgets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Layouts and Widgets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

rtui provides 35+ widgets and 9 layout containers. This guide covers every one
with practical examples you can copy and run.

> **Terminal only:** All examples in this guide must be saved as `.R` files
> and run from a real terminal with `Rscript my_app.R`. rtui apps do **not**
> work in RStudio, R GUI, Jupyter, or any embedded R console. See
> [Getting Started](getting-started.html) for setup instructions.

## How layouts work

Layouts are built by nesting container functions. Each container accepts child
widgets as arguments and arranges them according to its rules:

```{r}
library(rtui)

# Vertical stack with a horizontal row inside
quick_app(
  title = "Layout Demo",
  layout = vstack(
    header(),
    text("Top section", id = "top"),
    hstack(
      text("Left", id = "left"),
      text("Right", id = "right")
    ),
    footer()
  ),
  bindings = list(binding("q", "quit_app", "Quit", priority = TRUE)),
  on_action = function(event, state) {
    if (event$value == "quit_app") return(quit())
    state
  }
)
```

Every widget and container accepts optional `id` and `classes` parameters.
The `id` is used for event handling and updates; `classes` are used for CSS
styling.

## Layout containers

### vstack() -- Vertical stacking

Children are stacked top to bottom. This is the most common container:

```{r}
vstack(
  text("Line 1"),
  text("Line 2"),
  text("Line 3"),
  id = "my_stack"
)
```

### hstack() -- Horizontal stacking

Children are placed side by side, left to right:

```{r}
hstack(
  button("Save", id = "save"),
  button("Cancel", id = "cancel"),
  button("Help", id = "help")
)
```

### grid() -- CSS grid layout

Arrange children in a grid with specified rows and columns:

```{r}
grid(
  text("Cell 1"), text("Cell 2"),
  text("Cell 3"), text("Cell 4"),
  cols = 2,
  id = "my_grid"
)
```

You can set `rows`, `cols`, or both. Use CSS for fine-grained control over
column/row sizes:

```{r}
quick_app(
  layout = vstack(
    grid(
      text("Sidebar"), text("Main content"),
      cols = 2, id = "layout"
    )
  ),
  css = "
    #layout {
      grid-size: 2;
      grid-columns: 1fr 3fr;
      height: 1fr;
    }
  "
)
```

### center() and middle() -- Alignment

`center()` centres children horizontally; `middle()` centres vertically.
Combine them for perfect centring:

```{r}
center(
  middle(
    text("I'm in the exact centre!")
  )
)
```

### scroll() -- Scrollable container

Wraps content in a scrollable viewport:

```{r}
scroll(
  text("Line 1"),
  text("Line 2"),
  # ... many more lines
  text("Line 100"),
  id = "scroller"
)
```

### box() -- Bordered container

Wraps a single child with an optional border and title:

```{r}
box(
  text("Important content"),
  border = "round",
  title = "Notice"
)
```

Border styles: `"none"`, `"round"`, `"heavy"`, `"double"`.

### collapsible() -- Expandable section

A section that can be collapsed/expanded by clicking:

```{r}
collapsible(
  "Advanced Settings",
  input(placeholder = "API endpoint", id = "endpoint"),
  checkbox("Enable debug mode", id = "debug"),
  collapsed = TRUE,
  id = "advanced"
)
```

### tabs() and tab_pane() -- Tabbed content

Organise content into switchable tabs:

```{r}
tabs(
  tab_pane("Overview",
    text("Dashboard overview here"),
    title = "Overview", id = "tab1"
  ),
  tab_pane("Details",
    text("Detailed information here"),
    title = "Details", id = "tab2"
  ),
  tab_pane("Settings",
    text("App settings here"),
    title = "Settings", id = "tab3"
  ),
  id = "main_tabs"
)
```

### container() -- Plain block

A minimal container with no special layout behaviour. Useful as a CSS
styling target:

```{r}
container(
  text("Styled content"),
  id = "panel",
  classes = "highlight"
)
```

## Display widgets

### text() -- Plain text

Displays a text string. Updateable via `content`:

```{r}
text("Hello world", id = "greeting")

# Update later
update(state$app, "greeting", content = "Goodbye world")
```

### static() -- Rich text

Like `text()` but supports
[Rich markup](https://rich.readthedocs.io/en/latest/markup.html) for colours,
bold, italic, etc.:

```{r}
static("[bold red]Error:[/bold red] Something went wrong", id = "msg")
static("[green]Status:[/green] All systems operational", id = "status")
```

### markdown() -- Rendered Markdown

Renders Markdown content with headings, lists, code blocks, and more:

```{r}
markdown("# Welcome\n\nThis is **markdown** content.\n\n- Item 1\n- Item 2",
         id = "docs")

# Update content dynamically
update(state$app, "docs", content = "# Updated\n\nNew content here.")
```

### digits() -- Large number display

Shows text in a large, blocky font -- ideal for counters, clocks, and
dashboards:

```{r}
digits("12:34", id = "clock")
digits("0", id = "counter")

# Update the display
update(state$app, "clock", value = "12:35")
```

Supports digits 0-9, colons, spaces, and periods.

### pretty_table() -- Rich-formatted table

Renders a data.frame as a formatted table (non-interactive):

```{r}
pretty_table(
  head(mtcars, 5),
  title = "Motor Trend Cars",
  id = "info_table"
)
```

### sparkline() -- Inline mini chart

A compact line chart for showing trends:

```{r}
sparkline(c(1, 4, 2, 8, 5, 3, 9, 6), id = "trend")

# Update with new data
update(state$app, "trend", data = c(3, 7, 2, 5, 8, 1, 4))
```

### progress_bar() -- Progress indicator

```{r}
progress_bar(total = 100, progress = 0, id = "pb")

# Update progress from a handler
update(state$app, "pb", progress = 42)
```

Options: `show_eta` (estimated time), `show_percentage`.

### rule() -- Horizontal divider

A horizontal line, optionally with a centred label:

```{r}
rule()                          # plain line
rule(label = "Section Break")   # labelled line
```

### loading() -- Loading spinner

An animated loading indicator:

```{r}
loading(id = "spinner")

# Show/hide it
update(state$app, "spinner", display = TRUE)
update(state$app, "spinner", display = FALSE)
```

### placeholder() -- Placeholder area

A labelled placeholder useful during development:

```{r}
placeholder("Chart goes here", id = "ph")
```

### log_view() -- Scrolling log output

An append-only scrolling log, ideal for status messages and debug output:

```{r}
log_view(id = "logs", max_lines = 500)

# Append lines from a handler
log_write(state$app, "logs", "Server started on port 8080")
log_write(state$app, "logs",
          "[bold green]OK[/bold green] Connected",
          markup = TRUE)
```

## Input widgets

### button() -- Clickable button

```{r}
button("Click Me", id = "btn", tooltip = "Does something cool")

# Handle clicks
on_click = list(
  btn = function(event, state) {
    notify(state$app, "Button clicked!")
    state
  }
)
```

Update the label or disable it at runtime:

```{r}
update(state$app, "btn", label = "Clicked!")
update(state$app, "btn", disabled = TRUE)
```

### input() -- Single-line text input

```{r}
input(
  placeholder = "Enter your name...",
  value = "",
  id = "name_input",
  tooltip = "Type here",
  validators = "number"  # optional validation
)
```

Validators: `"number"`, `"integer"`, `"url"`, or `"regex:PATTERN"`.

Handle value changes and submission (Enter key):

```{r}
on_change = list(
  name_input = function(event, state) {
    # event$value has the current text
    state
  }
),
on_submit = list(
  name_input = function(event, state) {
    notify(state$app, paste("Hello,", event$value))
    state
  }
)
```

### text_area() -- Multi-line editor

```{r}
text_area(
  value = "Edit me...",
  language = "r",            # syntax highlighting
  show_line_numbers = TRUE,
  id = "editor"
)
```

The `language` parameter enables syntax highlighting. Supported languages
include `"r"`, `"python"`, `"javascript"`, `"markdown"`, `"json"`, `"sql"`,
and many more.

### masked_input() -- Formatted input

Input with a fixed template for structured data:

```{r}
masked_input(template = "999-999-9999", id = "phone")
masked_input(template = "AA99 9AA", id = "postcode")
```

Template characters: `A` (letter), `9` (digit), `!` (force uppercase),
`>` (force uppercase following), `<` (force lowercase following). All other
characters are literal separators.

### checkbox() -- Toggle with label

```{r}
checkbox("Enable notifications", value = TRUE, id = "notif")

on_change = list(
  notif = function(event, state) {
    # event$value is TRUE or FALSE
    state
  }
)
```

### switch_input() -- Toggle switch

```{r}
switch_input(value = FALSE, id = "dark_mode")
```

### select() -- Dropdown menu

```{r}
# Simple options
select(c("Small", "Medium", "Large"), id = "size")

# Named options (display => value)
select(
  c("Small (S)" = "s", "Medium (M)" = "m", "Large (L)" = "l"),
  prompt = "Pick a size...",
  id = "size"
)
```

### radio_set() and radio_button() -- Radio group

```{r}
radio_set(
  radio_button("Option A"),
  radio_button("Option B"),
  radio_button("Option C", value = TRUE),  # pre-selected
  id = "choice"
)
```

### data_table() -- Interactive data table

A full-featured interactive table with sorting and row selection:

```{r}
data_table(
  mtcars,
  id = "cars_table",
  cursor = "row",          # "cell", "row", "column", "none"
  zebra_stripes = TRUE,
  sortable = TRUE
)
```

Update at runtime -- add rows, clear data:

```{r}
# Add rows (pass a data.frame as a list)
update(state$app, "cars_table", add_rows = as.list(new_df))

# Clear all data
update(state$app, "cars_table", clear_data = TRUE)
```

### option_list() -- Selectable list

A scrollable list of options. Fires change events on selection:

```{r}
option_list(
  items = c("Apple", "Banana", "Cherry", "Date"),
  id = "fruits"
)

# Update items dynamically
update(state$app, "fruits", items = c("Fig", "Grape", "Kiwi"))
```

### selection_list() -- Multi-select list

Like `option_list()` but allows multiple selections:

```{r}
selection_list(
  items = c("Read", "Write", "Execute"),
  id = "permissions"
)
```

### list_view() -- General purpose list

```{r}
list_view(
  items = c("Item 1", "Item 2", "Item 3"),
  id = "my_list"
)
```

## Navigation widgets

### header() and footer()

The header displays the app title; the footer shows key bindings:

```{r}
vstack(
  header(show_clock = TRUE),
  text("Content"),
  footer()
)
```

### tree() -- Custom tree view

Display hierarchical data:

```{r}
tree(
  label = "Project",
  data = list(
    src = list("main.R", "utils.R", "app.R"),
    tests = list("test-main.R", "test-utils.R"),
    "README.md"
  ),
  id = "project_tree"
)
```

### directory_tree() -- File system browser

Browse a real directory:

```{r}
directory_tree(path = ".", id = "files")
```

Fires click events with the selected file path:

```{r}
on_click = list(
  files = function(event, state) {
    path <- event$value$path
    notify(state$app, paste("Selected:", path))
    state
  }
)
```

### content_switcher() -- Show one child at a time

```{r}
content_switcher(
  text("Page 1 content", id = "page1"),
  text("Page 2 content", id = "page2"),
  initial = "page1",
  id = "switcher"
)
```

## Common widget properties

Most widgets support these runtime updates:

```{r}
# Show / hide a widget
update(state$app, "widget_id", display = TRUE)
update(state$app, "widget_id", display = FALSE)

# Enable / disable a widget
update(state$app, "widget_id", disabled = TRUE)
update(state$app, "widget_id", disabled = FALSE)

# Change text content (text, static, markdown)
update(state$app, "widget_id", content = "New text")

# Change value (input, digits, switch)
update(state$app, "widget_id", value = "new value")

# Change button/checkbox label
update(state$app, "widget_id", label = "New Label")
```

## CSS styling

rtui supports Textual's CSS dialect for fine-grained styling. Pass a CSS
string to the `css` parameter:

```{r}
quick_app(
  layout = vstack(
    text("Styled!", id = "styled_text"),
    id = "root"
  ),
  css = "
    #styled_text {
      color: cyan;
      text-style: bold;
      text-align: center;
      padding: 2;
      border: heavy green;
    }
    #root {
      align: center middle;
    }
  "
)
```

Common CSS properties:

- `background`, `color` -- colours
- `width`, `height` -- sizes (`auto`, `1fr`, `50%`, `30`, etc.)
- `padding`, `margin` -- spacing
- `border` -- `none`, `tall`, `heavy`, `round`, `double`
- `text-align` -- `left`, `center`, `right`
- `text-style` -- `bold`, `italic`, `underline`, `reverse`
- `display` -- `block`, `none`
- `dock` -- `top`, `bottom`, `left`, `right`
- `align` -- `left top`, `center middle`, `right bottom`

Combine themes with your own CSS by concatenating:

```{r}
css = paste0(
  tui_theme("dracula"),
  "
  #sidebar { width: 30; border-right: tall $accent; }
  #main { width: 1fr; padding: 1; }
  "
)
```

The `$accent`, `$surface`, `$text`, and `$text-muted` variables from Textual
are available in your CSS.

## Tooltips

Many widgets support a `tooltip` parameter that shows hover text:

```{r}
button("Save", id = "save", tooltip = "Save the current document")
input(placeholder = "Search...", id = "search", tooltip = "Type to filter")
```

## Example: Dashboard layout

Here's a complete dashboard layout combining multiple concepts:

```{r}
library(rtui)

quick_app(
  title = "Dashboard",
  layout = vstack(
    header(),
    hstack(
      # Sidebar
      vstack(
        static("[bold]Navigation[/bold]"),
        rule(),
        option_list(
          items = c("Overview", "Reports", "Settings"),
          id = "nav"
        ),
        id = "sidebar"
      ),
      # Main content
      vstack(
        static("[bold]Welcome to the Dashboard[/bold]", id = "title"),
        rule(),
        hstack(
          box(digits("42", id = "metric1"), border = "round", title = "Users"),
          box(digits("128", id = "metric2"), border = "round", title = "Sales"),
          box(digits("99%", id = "metric3"), border = "round", title = "Uptime")
        ),
        text_plot(id = "chart"),
        id = "main"
      ),
      id = "content"
    ),
    footer()
  ),
  on_mount = function(event, state) {
    plot_bar(state$app, "chart",
             labels = c("Mon", "Tue", "Wed", "Thu", "Fri"),
             values = c(120, 200, 150, 180, 220),
             title = "Weekly Activity",
             color = "cyan")
    state
  },
  bindings = list(
    binding("q", "quit_app", "Quit", priority = TRUE)
  ),
  on_action = function(event, state) {
    if (event$value == "quit_app") return(quit())
    state
  },
  css = paste0(
    tui_theme("catppuccin"),
    "
    #sidebar { width: 25; padding: 1; border-right: tall $accent; }
    #main { width: 1fr; padding: 1; }
    #chart { height: 1fr; }
    "
  )
)
```
